Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-22 paper.tex
Some text

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes
(omitting values < 0.01\%).
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
/variant\_position & int32 & 111.6M & 2.0 & \\
/variant\_quality & float32 & 87.4M & 2.6 & \\
/variant\_allele & string & 69.3M & 13.0 & \\
% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
/variant\_AN & int32 & 47.3M & 4.8 & \\
% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
/variant\_filter & bool & 6.4M & 570.0 & \\
% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
/sample\_id & str & 268.12K & 2.3 & \\
% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
%%%% OLD for chr20
% call\_AD &  int16 & 179.7G & 26 & 24.0\\
% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
% call\_DPF& int16  & 115.1G & 20 & 15.3\\
% call\_FT &  string & 58.5G & 160 & 7.8 \\
% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
% variant\_allele & string & 21.5M & 11 \\
% variant\_position & int32 & 12.7M & 4.7 \\
% variant\_ABratio & float32 & 9.6M & 6.3 \\
% variant\_AN & int32 & 9.6M & 6.3 \\
% variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

@@

Once the VCF data has been converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. We take advantage of Zarr's flexibility
in \texttt{vcf2zarr} by choosing integer widths based

and reasonable compressor defaults, but these can be
modified by generating and editing a JSON-formatted storage schema.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from ICF and flushing to file. Similar to the \texttt{explode} command,
encoding to Zarr can be done in parallel on a single 
machine using the \texttt{encode} command,
or can be distributed across a cluster using the 
\texttt{dencode-init},
\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
The distributed commands are fault-tolerant, reporting any failed
parititions so that they can be retried.

\subsection{Choosing default compressor settings}
To inform the choice of compression settings across different fields 
in VCF data, we analysed their effect on compression ratio on 
data from the 1000 Genomes project.
[TODO, reffing Fig~\ref{fig-compression-shuffle} and 
Fig~\ref{fig-compression-chunksize}.]

% [FILL IN WITH Shadi's analysis from here:
% https://github.com/sgkit-dev/bio2zarr/discussions/74
% Discussed in:
% https://github.com/sgkit-dev/vcf-zarr-publication/issues/26

@@

@@

\begin{figure}
\includegraphics{figures/compression-shuffle}
\caption{Effects of shuffle setting on compression ratios on 1000 Genomes
data. TODO MORE DETAIL. 
\label{fig-compression-shuffle}}
\end{figure}

\begin{figure}
\includegraphics{figures/compression-chunksize}
\caption{Effects of chunk sizes on compression ratios on 1000 Genomes
data. TODO MORE DETAIL. 
\label{fig-compression-chunksize}}
\end{figure}

>>>>

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
Genomics England aggV2 VCFs on chromosome 2 using \texttt{vcf2zarr}
default settings. Each field is stored independently 
as a Zarr array with the given type (sufficient to represent all values in the
data). We show the total storage consumed (reported via \texttt{du}) in 
power-of-two units, and the compression ratio achieved on that array.
We also show the percentage of the overall storage that each array consumes
(omitting values < 0.01\%).
\label{tab-genomics-england-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
\midrule
/call\_AD & int16 & 658.4G & 26.0 & 25.35\% \\
/call\_GQ & int16 & 654.5G & 13.0 & 25.20\% \\
/call\_DP & int16 & 570.0G & 15.0 & 21.95\% \\
/call\_DPF & int16 & 447.1G & 20.0 & 17.22\% \\
/call\_PL & int16 & 162.6G & 160.0 & 6.26\% \\
/call\_GQX & int16 & 41.0G & 210.0 & 1.58\% \\
/call\_FT & string & 25.0G & 1400.0 & 0.96\% \\
/call\_genotype & int8 & 21.5G & 410.0 & 0.83\% \\
/call\_genotype\_mask & bool & 12.8G & 680.0 & 0.49\% \\
/call\_genotype\_phased & bool & 2.4G & 1900.0 & 0.09\% \\
/call\_PS & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADF & int8 & 383.4M & 12000.0 & 0.01\% \\
% /call\_ADR & int8 & 383.4M & 12000.0 & 0.01\% \\
/variant\_position & int32 & 111.6M & 2.0 & \\
/variant\_quality & float32 & 87.4M & 2.6 & \\
/variant\_allele & string & 69.3M & 13.0 & \\
% /variant\_OLD\_MULTIALLELIC & object & 55.88M & 8.2 & 0.00\% \\
/variant\_AN & int32 & 47.3M & 4.8 & \\
% /variant\_AC & int32 & 47M & 4.9 & 0.00\% \\
% /variant\_AC\_Het & int32 & 46.62M & 4.9 & 0.00\% \\
% /variant\_ABratio & float32 & 20.23M & 11.0 & 0.00\% \\
% /variant\_MendelSite & float32 & 20.09M & 11.0 & 0.00\% \\
% /variant\_medianGQ & float32 & 19.16M & 12.0 & 0.00\% \\
% /variant\_completeGTRatio & float32 & 14.98M & 15.0 & 0.00\% \\
% /variant\_phwe\_eur & float32 & 14.46M & 16.0 & 0.00\% \\
% /variant\_phwe\_afr & float32 & 13.37M & 17.0 & 0.00\% \\
% /variant\_medianDepthNonMiss & float32 & 12.17M & 19.0 & 0.00\% \\
% /variant\_medianDepthAll & float32 & 11.99M & 19.0 & 0.00\% \\
% /variant\_phwe\_sas & float32 & 11.59M & 20.0 & 0.00\% \\
% /variant\_AC\_Hom & int32 & 11.12M & 21.0 & 0.00\% \\
% /variant\_missingness & float32 & 9.3M & 25.0 & 0.00\% \\
% /variant\_phwe\_eas & float32 & 7.12M & 32.0 & 0.00\% \\
/variant\_filter & bool & 6.4M & 570.0 & \\
% /variant\_AC\_Hemi & int32 & 849.43 KiB & 280.0 & 0.00\% \\
% /variant\_composite\_filter & bool & 433.38 KiB & 130.0 & 0.00\% \\
% /variant\_OLD\_CLUMPED & object & 384.18 KiB & 1200.0 & 0.00\% \\
% /variant\_id & object & 298.76 KiB & 1600.0 & 0.00\% \\
% /variant\_phwe\_amr & float32 & 269.52 KiB & 870.0 & 0.00\% \\
% /variant\_id\_mask & bool & 269.46 KiB & 220.0 & 0.00\% \\
/sample\_id & str & 268.1K & 2.3 & \\
% /variant\_contig & int16 & 257.77 KiB & 450.0 & 0.00\% \\
% /contig\_length & int64 & 3.79 KiB & 5.3 & 0.00\% \\
% /contig\_id & object & 2.3 KiB & 8.8 & 0.00\% \\
% /filter\_id & object & 808 bytes & 0.6 & 0.00\% \\
%%%% OLD for chr20
% call\_AD &  int16 & 179.7G & 26 & 24.0\\
% call\_GQ &  int16 & 171.8G & 13 & 23.0 \\
% call\_DP &  int16 & 141.8G & 16 & 18.9 \\
% call\_DPF& int16  & 115.1G & 20 & 15.3\\
% call\_FT &  string & 58.5G & 160 & 7.8 \\
% call\_PL &  int16 & 51.4G & 140 & 6.9 \\
% call\_GQX &  int16 & 12.1G & 190 & 1.6 \\
% call\_genotype & int8 & 6.1G & 380 & 0.8 \\
% call\_genotype\_mask & bool & 3.7G  & 630 & 0.5\\
% call\_genotype\_phased & bool & 692.5M  & 1700 & 0.1 \\
% call\_PS  & int8  & 102.2M & 12000 & 0.05 \\
% variant\_quality & float32 & 22.6M & 2.7 & <0.01 \\
% variant\_allele & string & 21.5M & 11 \\
% variant\_position & int32 & 12.7M & 4.7 \\
% variant\_ABratio & float32 & 9.6M & 6.3 \\
% variant\_AN & int32 & 9.6M & 6.3 \\
% variant\_filter & bool & 1.9M & 490 \\
\bottomrule
\end{tabular}
\end{table}

@@

Once the VCF data has been converted to the intermediate columnar format,
it can then be converted to Zarr using the \texttt{vcf2zarr encode}
command. By default we choose integer widths based on the maximum
and minimum values observed during conversion to ICF
along with reasonable compressor defaults (see next section), 
but these can be
modified by generating a JSON-formatted storage schema, 
which can be edited and supplied as an argument to \texttt{encode}.
Encoding a given column (for example, \texttt{call\_AD})
involves creating a buffer to hold a full variant-chunk of the 
array in question, and then sequentially filling this buffer with 
values read from ICF and flushing to file.
 Similar to the \texttt{explode} command,
encoding to Zarr can be done in parallel on a single 
machine using the \texttt{encode} command,
or can be distributed across a cluster using the 
\texttt{dencode-init},
\texttt{dencode-partition} and \texttt{dencode-finalise} commands.
The distributed commands are fault-tolerant, reporting any failed
partitions so that they can be retried.

\subsection{Choosing default compressor settings}
To inform the choice of compression settings across different fields 
in VCF data, we analysed their effect on compression ratio on 
data from the 1000 Genomes project. We began by downloading 
the high coverage WGS data for 2504 samples~\citep{byrska2022high}
for chromosome 22 and converted to Zarr using 
\texttt{vcf2zarr} with default settings. We then systematically 
examined the effects of varying chunk sizes and compressor settings
on the compression ratio for call-level fields. We excluded call\_PL
from this analysis as it requires conversion to a ``local alleles''
encoding~\citep{poterba2024scalable} to be efficient, 
which is planned for implementation in a future 
version of \texttt{vcf2zarr}.

Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
fairly weak influence on compression ratio. While larger 
sample chunk sizes lead to a slight increase in compression
ratio on the more compressable fields, the overall effect
on total storage used will be slight. Variant chunk size 
appears to have no effect on compression ratio. Thus, we 
do not expect the choice of chunk size to have a significant
impact on overall storage usage, and so choice should be
determined by other considerations such as expected data
access patterns.

[TODO, reffing Fig~\ref{fig-compression-shuffle} and 
Fig~\ref{fig-compression-chunksize}.]

@@

% Acking according to the request here:
% https://ftp.1000genomes.ebi.ac.uk/vol1/ftp/data_collections/1000G_2504_high_coverage/20190405_1000G_2504_high_cov_data_reuse_README.md
The 1000 Genomes data used to analyse the effects of storage settings
was generated at the New York Genome Center with funds provided by
NHGRI Grant 3UM1HG008901-03S1.

@@

\begin{figure}
\includegraphics{figures/compression-chunksize}
\caption{Effects of chunk sizes on compression ratio on call-level 
fields in 1000 Genomes data.
(A) Varying sample chunk size, holding variant chunk size fixed at 10,000.
(B) Varying variant chunk size, holding sample chunk size fixed at 1,000.
In all cases the zstd compressor with clevel=7 was used. Bit shuffle
was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-chunksize}}
\end{figure}

\begin{figure}
\includegraphics{figures/compression-shuffle}
\caption{Effects of Blosc shuffle settings on compression ratio on call-level 
fields in 1000 Genomes data. 
In all cases the zstd compressor with clevel=7 was used, with a variant
chunk size of 10,000 and sample chunk size of 1,000.
\label{fig-compression-shuffle}}
\end{figure}
